---
title: "PDM PS4"
output: html
editor: visual
format:
  html:
    embed-resources: true
---

# PDM Final Assignment

*Who’s Working? Demographic Analysis of Employment Trends in the US (2001-2023)*

[Link to Repo](https://github.com/czhang009/PDM_PS4.git)

## Introduction

```{r, echo=F, include=F}
library(knitr) 
library(readr)
library(dplyr)
library(haven)
library(tidyverse)
library(ggplot2)
library(broom)
library(kableExtra)
library(scales)
library(nnet)

# read data
ipums <- read_dta("ipums.dta.gz")

# find number of observations + column names
dim(ipums)
names(ipums)
```

Employment inequality has long held its place among the systemic barriers that plague society, barriers rooted in historical biases and deeply embedded social structures that continue to influence who is afforded access to paid work. While progress has been made in many developed economies and norms around gender roles and family structures have shifted, the notion that employment inequality, particularly in relation to gender (though it must be acknowledged that employment inequality manifests along characteristics beyond gender), has been resolved can certainly create harmful agendas. Patterns of labor force participation continue to reflect disparities shaped by factors such as caregiving responsibilities and discriminatory hiring practices. Understanding who participates in the workforce and who is systematically excluded from it remains a key component of understanding inequality and economic opportunity.

This brief investigation seeks to understand how employment status, specifically whether an individual is employed, unemployed, or not in the labor force, varies by gender, marital status, and educational attainment, and how these relationships have evolved over time. Rather than focusing on post-employment outcomes such as wages, this assignment centers on differences in access to employment itself.

### Data

The data used in this analysis is derived from the Integrated Public Use Microdata Series (IPUMS), specifically the American Community Survey (ACS) samples for the years 2001 through 2023. The ACS is a large-scale, nationally representative survey administered by the U.S. Census Bureau that captures individual-level characteristics, such as demographics and housing conditions. The unit of observation is therefore the individual.

For the purposes of this assignment and the consideration of the sheer size of each sample, I condensed the density of each annual ACS sample from its original 1% random sample to 0.01% and restricted the sample selection to only include those within the ages of 16 to 65 (working-age population). In the data extraction, there are 17 variables across 450,086 observations. Only 6 of the 17 variables are relevant to this assignment; the rest are pre-selected variables. I found the IPUMS codebook to be straightforward.

## Variables

```{r, echo=F, include=F}
# continuous vars descriptive stats
summary(ipums$age)
sd(ipums$age)

# recode categorical variables
ipums_rec <- ipums %>%
  mutate(empstat_rec = case_when(
    empstat %in% 1 ~ "Employed",
    empstat %in% 2 ~ "Unemployed",
    empstat %in% 3 ~ "Not in labor force")) %>%
  mutate(marst_rec = case_when(
    marst %in% c(1, 2) ~ "Married",
    marst %in% c(3, 4, 5) ~ "Previously married",
    marst %in% 6 ~ "Single")) %>%
  mutate(educ_rec = case_when(
    educ %in% c(0, 1, 2, 3, 4, 5) ~ "No high school",
    educ %in% c(6, 7, 8) ~ "High school completion",
    educ %in% c(10, 11) ~ "College completion")) %>%
  mutate(sex_rec = case_when(
    sex %in% 1 ~ "Male",
    sex %in% 2 ~ "Female"))

# categorical vars descriptive stats
# empstat
emp <- table(ipums_rec$empstat_rec)
emp
round(100*prop.table(emp), digits=0)
# marst
mar <- table(ipums_rec$marst_rec)
mar
round(100*prop.table(mar), digits=0)
# educ
edu <- table(ipums_rec$educ_rec)
edu
round(100*prop.table(edu), digits=0)
# sex
se <- table(ipums_rec$sex_rec)
se
round(100*prop.table(se), digits=0)
```

#### Continuous Variables: Descriptive Statistics

| Variable Name | Min | Max | Median | Mean  | Std   |
|---------------|-----|-----|--------|-------|-------|
| age           | 16  | 65  | 42     | 41.13 | 14.40 |

*Note that age has been pre-filtered on IPUMs to include only the working-age population (16-65).*

#### Categorical Variables: Descriptive Statistics

| Variable Name | Definition | Categories | Frequency | Sample % |
|---------------|---------------|---------------|---------------|---------------|
| empstat | Respondent's employment status | Employed (empstat = 1) | 306,355 | 68% |
|  |  | Unemployed (empstat = 2) | 21,136 | 5% |
|  |  | Not in labor force (empstat = 3) | 122,595 | 27% |
| marst | Respondent's current marital status | Married (marst = 1, 2) | 238,804 | 53% |
|  |  | Previously married (separated, divorced, or widowed; marst = 3, 4, 5) | 64,369 | 14% |
|  |  | Single (marst = 6) | 146,913 | 33% |
| educ | Respondent's highest level of educational attainment | No high school completion (educ = 0, 1, 2, 3, 4, 5) | 57,503 | 13% |
|  |  | High school completion (educ = 6, 7, 8) | 264,152 | 59% |
|  |  | College completion (educ = 10, 11) | 128,431 | 29% |
| sex | Respondent's reported sex | Male (sex = 1) | 221,928 | 49% |
|  |  | Female (sex = 2) | 228,158 | 51% |

## Analysis

```{r univariate, echo=F}
# univariate: empstat frequency dist
ggplot(ipums_rec, aes(x = empstat_rec)) +
  geom_bar(fill = "#5DA5DA") +
  labs(title = "Figure 1: Distribution of Employment Status",
       x = "Employment Status",
       y = "Count") +
  theme_minimal() +
  scale_y_continuous(labels = comma) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), # removes grid from background
        panel.spacing = unit(1, "lines"), 
        panel.background = element_rect(fill = "white")) # add unfilled rectangle to   border graph
```

Figure 1 shows the univariate distribution of employment status among individuals aged 16-65 in the IPUMS sample. The majority of individuals fall into the "Employed" category. A smaller yet still sizeable share are categorized as "Not in labor force". "Unemployed" is the smallest category.

```{r line, echo=F}
# line plot: employment status (colored by sex) across time

# manually set levels (order)
ipums_rec$empstat_rec <- factor(ipums_rec$empstat_rec, levels = c("Employed", "Unemployed", "Not in labor force"))

plot <- ipums_rec %>%
  group_by(year, sex_rec, empstat_rec) %>%
  summarise(count = n(), .groups = 'drop') %>%
  group_by(year, sex_rec) %>%
  mutate(prop = count/sum(count))

ggplot(plot, aes(x = year, y = prop, color = sex_rec)) +
  geom_line() +
  facet_wrap(~empstat_rec) +
  labs(title = "Figure 2: Employment Status by Sex Over Time (2001–2023)",
       x = "Year", y = "Proportion",
       color = "Sex") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(1, "lines"), 
        panel.background = element_rect(fill = "white"))
```

Figure 2 shows trends in employment status over time by sex (for the years 2001-2023). In the employed panel, men are consistently employed at a higher proportion than women are. The gap between men and women is more narrow for the unemployed category; the proportions appear to converge in recent years. Women have a consistently higher proportion in the "not in labor force" category.

```{r bar, echo=F}
# grouped bar chart: marital status vs employment status 
plot2 <- ipums_rec %>%
  group_by(marst_rec, empstat_rec) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(marst_rec) %>%
  mutate(prop = n/sum(n))

ggplot(plot2, aes(x = marst_rec, y = prop, fill = empstat_rec)) +
  geom_bar(stat = 'identity', position = 'dodge') +
   labs(title = "Figure 3: Employment Status by Marital Status",
       y = "Proportion", x = "Marital Status", fill = "Employment Status") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(1, "lines"), 
        panel.background = element_rect(fill = "white")) +
   scale_fill_manual(values = c("Employed" = "#C3B1E1",
                                "Unemployed" = "#A1C9F4",
                                "Not in labor force" = "#8DE5A1"))
```

Figure 3 shows how employment status varies across marital statuses. Married individuals are employed at a higher proportion than those previously married or single. The trend is reversed for the unemployed and "not in labor force" employment statuses: those married have the lowest proportion in the 2 employment categories, while single individuals have the highest proportions.

```{r bar1, echo=F}
# grouped bar chart: education vs employment status
plot3 <- ipums_rec %>%
  group_by(educ_rec, empstat_rec) %>%
  summarize(n = n(), .groups = "drop") %>%
  group_by(educ_rec) %>%
  mutate(prop = n/sum(n))

ggplot(plot3, aes(x = educ_rec, y = prop, fill = empstat_rec)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = "Figure 4: Employment Status by Education Level",
       y = "Proportion", x = "Education", fill = "Employment Status") +
  theme_minimal() + 
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(1, "lines"), 
        panel.background = element_rect(fill = "white")) +
  scale_fill_brewer(palette = "Set2")
```

Figure 4 displays how employment status varies by educational attainment. Those with college completion are employed at the highest proportion, while those with that did not complete high school are employed at the lowest proportion (out of the 3 education categories here). Those without a high school diploma have the highest proportion in the unemployed and "not in labor force" categories.

```{r line1, echo=F}
# line graph: employment status by age
plot4 <- ipums_rec %>%
  group_by(age, empstat_rec) %>%
  summarise(count = n(), .groups = 'drop') 

ggplot(plot4, aes(x = age, y = count, color = empstat_rec)) +
  geom_line() +
  labs(title = "Figure 5: Employment Status by Age",
       x = "Age", y = "Count",
       color = "Employment Status") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 14),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        panel.spacing = unit(1, "lines"), 
        panel.background = element_rect(fill = "white")) +
   scale_fill_viridis_d()
```

Figure 5 shows how employment status varies by age. Right around the age range of early 20s, the slope of the employed line rises dramatically, while the opposite trend is observed for the "not in labor force" category. The slope for the employed line continues to climb, though at a slower rate, until around the 50s, while the "not in labor force" line remains relatively flat until this age range. After this point, the "not in labor force" line begins to climb rapidly, while the employed line declines steeply. The unemployed line shows a slight peak just before the 20s, but otherwise the line remains relatively flat without much fluctuation.

### Regression

```{r, echo=F, include=T, message=F}
# scale year variable
ipums_rec$year_scaled <- scale(ipums_rec$year)

# supress message containing weights and iter lines
invisible(capture.output(
  model <- multinom(empstat_rec ~ sex_rec + marst_rec + educ_rec + year_scaled, data = ipums_rec)
))

# tidy model output + remove stat column
tidy_model <- tidy(model) %>%
  mutate(estimate = number(estimate, accuracy = 0.001),
         std.error = number(std.error, accuracy = 0.001),
         p.value = case_when(
      p.value < 0.001 ~ "< 0.001",  
      p.value >= 0.001 & p.value < 0.01 ~ paste("p =", round(p.value, 3)),  
      p.value >= 0.01 & p.value < 0.05 ~ paste("p =", round(p.value, 3)),  
      p.value >= 0.05 ~ paste("p =", round(p.value, 3)))) %>%
  select(-statistic)

# formatting table
tidy_model %>%
  kable(format = "html", digits = 3, caption = "Multinomial Logistic Regression:Employment Status") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = FALSE, position = "left")
```

The table above is the output from a multinominal logistic regression where the DV is employment status. The coefficients shown are log-odds of being in either the unemployed or "not in labor force" group relative to being employed (reference category).

For the unemployed status, men (sex_recMale) are less likely than women to be unemployed (coefficient = -0.053, p \< 0.001). This effect is statistically significant, though small in magnitude. For the "not in labor force" status, men are also less likely than women to fall under this category (coefficient = -0.533, p \< 0.001) - the magnitude on this effect is greater than that for the unemployed category.

Previously married individuals are more likely than married individuals to be unemployed than employed (coefficient = 0.563, p \< 0.001) and are also more likely to be "not in labor force" than employed (coefficient = 0.160, p \< 0.001). Single individuals are even more likely to be unemployed (coefficient = 0.995, p \< 0.001) and also even more likely to be out of the labor force (coefficient = 0.380, p \< 0.001) (relative to married individuals).

High school graduates are more likely than those with a bachelor's degree to be both unemployed (coefficient = 0.689, p \< 0.001) and out of labor force (coefficient = 0.759, p \< 0.001). Those that did not complete high school are even more likely to be unemployed (coefficient = 1.302, p \< 0.001) and out of labor force (coefficient = 1.934, p \< 0.001) (relative to those with a bachelor's degree).

Over time, unemployment appears to decrease (coefficient = -0.140, p \< 0.001), but the "not in labor force" status appears to increase (coefficient = 0.066, p \< 0.001).

## Citation

Ruggles, S., Flood, S., Sobek, M., Backman, D., Chen, A., Cooper, G., Richards, S., Rodgers, R., & Schouweiler, M. (2025). IPUMS USA: Version 16.0 American Community Survey, 2023 \[Data set\]. IPUMS. doi.org/10.18128/D010.V16.0
